Evaluating the Performance and Energy Efficiency of COSMO-ART, a Fully Online Coupled Model System with Numerical Weather Forecast and Chemical Transport Models

### 2014-04-04 paper_enahpc14/Introduction.tex
adds latest modifications to the document

Comprehensive  scientific assessments  of climate-warming  trends over
the  past 50  years have  shown  that the  state and  dynamics of  the
Earth's climate system have  undergone unprecedented major changes, as
evidenced  by  increases  in   global  average  atmosphere  and  ocean
temperatures,  widespread melting  of  snow and  ice,  and rising  sea
levels.   The last report  of the  Intergovernmental Panel  on Climate
Change, released  in September 2013  \citep{IPCC-2013}, concluded that
most of the observed temperature  increases since the mid 20th Century
have  been caused  by  increasing concentrations  of greenhouse  gases
resulting from  anthropogenic activities  such as fossil  fuel burning
and  deforestation.  Long-lived greenhouse  gases, for  example carbon
dioxide, methane  and nitrous oxide, are chemically  stable and linger
homogeneously well-mixed and over time scales of a decade to thousands
of years in  the atmosphere; whereas short-lived gases  such as water,
suphur  dioxide  and  carbon  monoxide  which  respond  physically  or
chemically  to changes  in temperature,  act primarily  as  a feedback
mechanism.   Some   extremly  powerful  greenhouse   gases,  such  as
fluorinated gases can even have  a warming effect on the atmosphere up
to  23000  times  greater  than  carbon  dioxide  \citep{Zehner-2012}.
Fortunately  the  European  Parliament,  which  has  gained  worldwide
recognition as a  leader in climate policy, has  passed legislation in
March 2014  to phase  out fluorinated gas  emissions by  two-thirds by
2030,  towards  the  internationally  agreed goal  of  keeping  global
warming  below  2  degrees  Celsius  compared to  the  temperature  in
pre-industrial times.

Current  cumulative releases of  human-induced gas  emissions enhances
the greenhouse  effect to  the Earth's atmosphere  and thus  creates a
large imbalance between incoming solar absorbed radiation and outgoing
longwave radiation emitted back  to space, through radiative trapping.
While the  Earth's temperature  is dependent upon  the greenhouse-like
action of  the atmosphere, the  Earth's radiation balance  is strongly
influenced by several  other factors such as the  type of surface that
sunlight first  encounters.  Forests, grasslands,  ocean surfaces, ice
caps, deserts,  and cities all absorb, reflect,  and radiate radiation
differently.   Sunlight falling  on a  white glacier  surface strongly
reflects back into space, resulting  in minimal heating of the surface
and  lower atmosphere.   Sunlight falling  on  a dark  desert soil  is
strongly absorbed,  on the other hand, and  contributes to significant
heating of the surface and lower atmosphere.  Cloud cover also affects
greenhouse  warming by  both reducing  the amount  of  solar radiation
reaching the earth's  surface and by reducing the  amount of radiation
energy emitted into space.

Aerosols are  suspensions of liquid,  solid or mixed particles  in the
air  (including sea  salt,  mineral dust,  sulphate, nitrate,  organic
carbon and  black carbon),  with highly variable  chemical composition
and  size  distribution \citep{Putaud-2010}.   Although  they are  not
considered a heat-trapping greenhouse gas and have shorter atmospheric
lifetimes, they significantly  affect the atmospheric radiative fluxes
and  are acknowledged  as one  of the  most significant  and uncertain
aspects    in    anthropogenic    forcing    over   the    last    150
years \citep{Koch-2009,  IPCC-2013}.  Enhanced aerosols concentrations
can impact the  climate system by reflecting (e.g.   pure sulfates and
nitrates)  or  absorbing  (e.g.   black carbon)  solar  radiation  and
thereby  exert a  cooling or  warming effect  on  the Earth-atmosphere
system,      causing       a      so-called      direct      radiative
forcing     \citep{Charlson-1991,    Haywood-2000,    Ramanathan-2001,
Liao-2005, Bangert-2012, Lundgren-2013}.   Depending on their size and
chemical composition, they can also  act as cloud condensation and ice
nuclei, and  thereby influence  the cloud microphysical  processes and
optical   properties  (cloud  albedo   effect,  \citealp{Twomey-1977})
through the contribution to  cloud formation.  This results in changes
in      droplet      concentrations     \citep{Albrecht-1989}      and
precipitation     \citep{Rosenfeld-2000,    Khain-2008,    Khain-2009,
Pruppacher-2010,   Seifert-2012,  Tao-2012,  Lee-2012},   producing  a
so-called  negative  indirect  radiative forcing  \citep{Haywood-2000,
Lohmann-2005,  VandenHeever-2011, Rosenfeld-2013}.  Wet  scavenging of
aerosol  particles  represents  a  major  removal  mechanism  for  air
pollutants from  the atmosphere, however the  effects of anthropogenic
aerosols on  clouds and  the hydrological cycle  as well as  the cloud
lifetime    effect    are    especially    hard    to    assess    and
quantify   \citep{IPCC-2013},   and   remain   one  of   the   largest
uncertainties in climate modeling and in climate change prediction due
to  the lack  of  knowledge in  cloud feedbacks  \citep{Sherwood-2013,
Rosenfeld-2013, Lee-2013}.

@@

COSMO-ART is  computationally much more demanding than  the COSMO core
since a  large number of additional  tracers and processes  have to be
considered.  Thus this model is currently severely-limited in terms of
applicability and expensive in terms of energy consumption.  COSMO has
recently  been  ported  to  GPUs  within the  framework  of  the  High
Performance  and  High  Productivity  Computing (HP2C)  Initiative  to
optimize it  for computational and energy  efficiency.  Although these
developments will  facilitate the  application of COSMO  for numerical
weather prediction and climate  simulations, they do little to address
the  coupling with  the  ART model  extension,  for which  significant
investments  are still required  to take  it to  a similar  level. The
efficiency of ART is being  addressed in the EU Exa2Green project. The
ultimate goal  of the  project is to  deliver a prototype  code, which
provides an energy  efficiency of at least five  times of the baseline
value.   Such   an  implementation   would  allow  the   community  to
investigate critical  questions at  higher resolution and  over longer
periods, at reduced cost  to the environment.  In Sec.~\ref{sec:1}, we
introduce the regional  atmospheric model COSMO-ART \citep{Vogel-2009}
that  accounts for feedbacks  between chemistry,  aerosols, radiation,
and clouds and additionally specify its technical setup related to the
investigated    performance    and     energy    evaluation    methods.
Sec.~\ref{sec:2}  describes the different  HPC systems  and associated
power  measurement  equipments used  for  conducting the  benchmarking
study and  presents energy-aware techniques employed  to determine the
energy  footprint  of  the considered  application.   Sec.~\ref{sec:3}
gathers benchmark  results from  the Exa2Green project  and highlights
areas where improvements will be necessary for the subsequent baseline
refactoring.  Some related work  will be discussed in Sec.~\ref{sec:4}
Finally, we  conclude with practical guidelines  to identifying energy
saving opportunities  and give directions for  further developments of
the modeling system in Sec.~\ref{concl}.

>>>>

Comprehensive  scientific assessments  of climate-warming  trends over
the  past 50  years have  shown  that the  state and  dynamics of  the
Earth's climate system have  undergone unprecedented major changes, as
evidenced  by  increases  in   global  average  atmosphere  and  ocean
temperatures,  widespread melting  of  snow and  ice,  and rising  sea
levels.   The last report  of the  Intergovernmental Panel  on Climate
Change, released  in September 2013  \citep{IPCC-2013}, concluded that
most of the observed temperature  increases since the mid 20th Century
have  been caused  by  increasing concentrations  of greenhouse  gases
resulting from  anthropogenic activities  such as fossil  fuel burning
and  deforestation.

The Earth's  surface temperature is determined by  the balance between
incoming  solar radiation and  outgoing infrared  radiation. Radiative
forcing describes any imbalance  in Earth's radiation budget caused by
human  or natural interventions  in the  climate system.   It directly
measures the capacity of a gas  or other forcing agents to affect that
energy balance, thereby contributing to climate change.  The intensity
of  solar energy,  reflectivity  or absorption  of  gases or  surfaces
(forests, grasslands, oceans, deserts, cloud cover, ice caps, cities),
emission of heat by various  materials, all differ in their warming or
cooling contributions to climate forcing.

The radiative forcing  potential of a greenhouse gas  is determined by
its  atmospheric concentration, warming  capacity, residence  time and
spatial  distribution.   Long-lived greenhouse  gases  such as  carbon
dioxide, methane and nitrous oxide, which linger chemically stable and
homogeneously well-mixed over time scales  of a decade to thousands of
years  in the  atmosphere,  are major  human-induced climate  drivers;
whereas  short-lived gases such  as water,  suphur dioxide  and carbon
monoxide,  which are  abundant heat-trapping  gases but  physically or
chemically  reactive  to changes  in  temperature,  cause positive  or
negative  feedbacks effects which  may vary  with forcing  history and
climate  state. On a  per molecule  basis, other  gases can  have more
potent  heat-trapping ability  than carbon  dioxide  (e.g. fluorinated
gases  are 23000  times more  effective \citep{Zehner-2012}),  but are
fortunately far less abundant in the atmosphere.

One  factor   overwhelmingly  affects  the   uncertainty  in  defining
anthropogenic  radiative  forcing:  the  effects  of  aerosols  (small
airborne liquid,  solid or mixed  particles, \citealp{Putaud-2010}) in
the  atmosphere.  Although  they  are not  considered a  heat-trapping
greenhouse   gas  and   have  shorter   atmospheric   lifetimes,  they
significantly affected the atmospheric  radiative fluxes over the last
150    years   \citep{Koch-2009,   IPCC-2013}.     Enhanced   aerosols
concentrations can impact the climate system by reflecting (e.g.  pure
sulfates  and  nitrates)  or  absorbing  (e.g.   black  carbon)  solar
radiation  and  thereby exert  a  cooling  or  warming effect  on  the
Earth-atmosphere   system,  causing   a  so-called   direct  radiative
forcing     \citep{Charlson-1991,    Haywood-2000,    Ramanathan-2001,
Liao-2005,  Bangert-2012,  Lundgren-2013}.   Depending on  their  size
distribution  and chemical  composition, they  can also  act  as cloud
condensation  and   ice  nuclei,  and  thereby   influence  the  cloud
microphysical   processes  and   optical   properties  (cloud   albedo
effect,  \citealp{Twomey-1977})  through  the  contribution  to  cloud
formation.      This     results     in     changes     in     droplet
concentrations                \citep{Albrecht-1989}                and
precipitation     \citep{Rosenfeld-2000,    Khain-2008,    Khain-2009,
Pruppacher-2010,   Seifert-2012,  Tao-2012,  Lee-2012},   producing  a
so-called  negative  indirect  radiative forcing  \citep{Haywood-2000,
Lohmann-2005,  VandenHeever-2011, Rosenfeld-2013}.  Wet  scavenging of
aerosol  particles  represents  a  major  removal  mechanism  for  air
pollutants from  the atmosphere, however the  effects of anthropogenic
aerosols on  clouds and  the hydrological cycle  as well as  the cloud
lifetime    effect    are    especially    hard    to    assess    and
quantify   \citep{IPCC-2013},   and   remain   one  of   the   largest
uncertainties in climate modeling and in climate change prediction due
to  the lack  of  knowledge in  cloud feedbacks  \citep{Sherwood-2013,
Rosenfeld-2013, Lee-2013}.

@@

COSMO-ART is  computationally much more demanding than  the COSMO core
since a  large number of additional  tracers and processes  have to be
considered.  Thus this model is currently severely-limited in terms of
applicability and expensive in terms of energy consumption.  COSMO has
recently  been  ported  to  GPUs  within the  framework  of  the  High
Performance  and  High  Productivity  Computing (HP2C)  Initiative  to
optimize it  for computational and energy  efficiency.  Although these
developments will  facilitate the  application of COSMO  for numerical
weather prediction and climate  simulations, they do little to address
the  coupling with  the  ART model  extension,  for which  significant
investments  are still required  to take  it to  a similar  level. The
efficiency of ART is being  addressed in the EU Exa2Green project. The
ultimate goal  of the  project is to  deliver a prototype  code, which
provides an energy  efficiency of at least five  times of the baseline
value.   Such   an  implementation   would  allow  the   community  to
investigate critical  questions at  higher resolution and  over longer
periods, at reduced cost  to the environment.

In  Sec.~\ref{sec:1},  we  introduce  the regional  atmospheric  model
COSMO-ART  \citep{Vogel-2009}  that  accounts  for  feedbacks  between
chemistry,  aerosols, radiation, and  clouds and  additionally specify
its technical setup related to the investigated performance and energy
evaluation  methods.   Sec.~\ref{sec:2}  describes the  different  HPC
systems   and  associated  power   measurement  equipments   used  for
conducting the benchmarking study and presents energy-aware techniques
employed  to   determine  the  energy  footprint   of  the  considered
application.   Sec.~\ref{sec:3}  gathers  benchmark results  from  the
Exa2Green  project and  highlights  areas where  improvements will  be
necessary for the subsequent  baseline refactoring.  Some related work
will  be  discussed  in  Sec.~\ref{sec:4} Finally,  we  conclude  with
practical  guidelines to identifying  energy saving  opportunities and
give  directions for further  developments of  the modeling  system in
Sec.~\ref{concl}.

### 2014-04-04 paper_enahpc14/Sec2:Power-measurement.tex
adds latest modifications to the document

\subsection{Hardware description}
\label{subsec:3.1}

\subsubsection{Monch (CSCS - ETH Zurich)}
While hardware  platforms mature  and are replaced,  we have  chosen a
state-of-the-art  dual-socket Intel  Ivybridge-based  platform (called
``Monch'')  for our power  measurements, as  it is  slated to  stay in
service without upgrade for the duration of the Exa2Green project.  In
principle at least,  this architecture could be recreated  or found in
an identical configuration beyond  the lifetime of the project.  Given
that  the baseline  benchmark  can be  reproduced  within an  expected
variance   (cf  previous   section),   and  that   the  baseline   run
configuration can be  used in all future versions of  the code, a fair
comparison  will  be  made  between  the baseline  and  the  milestone
versions of COSMO-ART.

@@

A snapshot of the code, which includes, at least conceptually, all the
information needed  to reproduce the  energy-to-solution benchmarks of
COSMO-ART, was  produced and run  on a full  rack of 52  nodes without
hyperthreading (hyperthreading was attempted, but always led to higher
energy-to-solution), i.e. on a total  of 1040 cores using 20 tasks per
node.

\subsection{Power-Performance measurement framework}
\label{subsec:3.2}

>>>>

\subsection{Hardware description}
\label{subsec:2.1}

\subsubsection{Monch (CSCS - ETH Zurich)}
While hardware  platforms mature  and are replaced,  we have  chosen a
state-of-the-art  dual-socket Intel  Ivybridge-based  platform (called
``Monch'')  for our power  measurements, as  it is  slated to  stay in
service without upgrade for the duration of the Exa2Green project.  In
principle at least,  this architecture could be recreated  or found in
an identical configuration beyond  the lifetime of the project.  Given
that  the baseline  benchmark  can be  reproduced  within an  expected
variance (see subsection~\ref{subsec:1.2}),  and that the baseline run
configuration can be  used in all future versions of  the code, a fair
comparison  will  be  made  between  the baseline  and  the  milestone
versions of COSMO-ART.

@@

\subsection{Power-Performance measurement framework}
\label{subsec:2.2}
